---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# MM_index_validation

These are the scripts for the article by Velek, Splinter et all, _Counting co-occurring diseases to predict mortality is as accurate as multimorbidity indices: results from an external validation study in the general population_

NB: The scripts rely on data that are not available in this repository. We are unable to place data in a public repository due to legal and ethical restraints. Sharing of individual participant data was not included in the informed consent of the study, and there is potential risk of revealing participants’ identities as it is not possible to completely anonymise the data.


All questions regarding the scripts should be directed to Premysl Velek, p.velek@erasmusmc.nl

Below is explanation of how the script was built and the workflow used. For details on specific analytic choices, refer to the respective functions in the `99_get_cohorts.R` script. Note that data cleaning is not included in the workflow as it builds on [our previous work](https://doi.org/10.1186/s12916-022-02487-x). Details can be found in `01_clean_index.R`, and in the specific scripts to clean data according to Robusto, 2016 and von Korff, 1992 (to be found in their respective folders).


```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction

This is the basic workflow for building specific cohort (subset of the Rotterdam Study
participants) with the following characteristics defined by the user:

* Follow-up for specific diseases (from among those available to us)
* Specific follow-up time
* Specific mean age at baseline
* Specific sex ratio (men only, women only or both)

The workflow has several functions (detailed below), the output are two data frames with data for the same 
cohort (ie. same participants, same follow-up time, the same baseline date and the same outcome (mortality)), but with different independent variables: 

* *index cohort* contains indicator variable for presence of set of user defined diseases at baseline
* *count cohort* contains the number of co-occurring diseases at baseline from among 10 non-communicable diseases with high burden in older populations (cancer, dementia, diabetes, coronary heart disease, heart failure, COPD, asthma, parkinsonism, stroke, depression) 

The goal is to compare the performance of two approaches to defining multimorbidity  (with respect to mortality): one is simple count of co-occurring diseases, the other is weighted approach in which different diseases have different weights. The weights are derived for different diseases coming from different multimorbidity indices.

## Data preparation

As the workflow builds specific cohort, it has to be repeated for each individual index you want to test. 

The workflow relies on a couple of libraries and a couple of helper functions. They all have to be loaded for it to work. All code in the document assumes you're working in an RStudio project. 

```{r preliminaries, echo=TRUE, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)

source(here::here("R", "get_cohorts.R"))

```

### Compile index

You first compile the index from follow-up data on specific diseases, defined by the user. The code below compiles the index by Tooth, 2008 which uses six diseases with the following weights:

* heart disease = 1
* stroke = 2
* low iron = 1
* lung disease = 2
* diabetes = 1
* cancer = 3
* alzheimer disease = 4


```{r compile, echo=TRUE, message = FALSE, warning = FALSE}
d <- c("hd", "stroke", "anemia", "lung", "dia", "can", "dem")

w <- c(1, 2, 1, 2, 1, 3, 4)

tooth <- compile_cohort(diseases = d)
```


### Index cohort

The function `get_index_cohort` takes as an input the index data, selects ERGO participants who have a complete follow up for particular diseases used in the given index and selects the baseline date (the follow up start) so that the mean age of the cohort at baseline matches with the mean age of the cohort used to develop that particular index. It can also select men or women only if that's the case of the original index. 

The arguments of the function are:

* index_data: data containing all the diseases in a given index. The following data have to be included:
  + Start and end of follow up for each individual disease and the prevalence/incidence indicators. The names of the variables have to be in the standard SHIFT form, e.g. 'startd_chd', 'endd_chd', 'inc_chd', 'prev_chd'. (Diseases that don't require any additional change to match the disease definition in the index cohort can be taken directly form the SHIFT data.) This structure have to be included even for additional variables for which we don't have longitudinal data (e.g. ADL). In this case, all cases will be prevalent and no case will be incident. The follow up start would then be the date of the interview, test, screening, etc. (In this particular case the end of follow up is not relevant and can stay empty) 
  + Information about the individual ERGO cohorts (RS-I, RS-II, RS-II).  
  + Date of birth (birthd), censor date (fp_censordate) and vital status for each participants (died), coded with 1 for death and 0 for living.
  + Overall follow-up start (the latest follow-up start among the individual diseases, fu_start) and follow_up end (the earliest follow up end from among the individual diseases, fu_end)

* longitudinal diseases: character vector with the diseases included in the index you want to validate. They have to be in the standard SHIFT form, as indicated in the list below. Importantly, as the name suggests, the diseases have to have continuous follow up.

  + dia: diabetes
  + stroke: stroke
  + can1: cancer, all types
  + dem: dementia
  + hf: heart failure
  + chd: coronary heart disease
  + park: parkinsonism
  + dep1: depression
  + COPD: chronic obstructive pulmonary disease
  + asthma: asthma
  
* cross_sectional_diseases: diseases, risk factors and lifestyle data measured only at one (or several) point in time.
* mean_age_index: the mean age of the index cohort, i.e. the mean age to which we need to get as close as possible to warrant fair comparison. Only integer values are allowed. **IMPORTANT NOTE: if your data does not allow you to move the follow up start (typically because you only have cross-sectional data for some diseases), ignore this argument. The function will not move the follow-up start.**
* sex_ratio_index: whether men only, women only or both men and women should be
 included in the cohort, based on the original cohort of each index. Only three
 values are allowed: women_only will results in a cohort with only women, men_only
 will result in a cohort with only men, both - the default - will not change the
 sex ratio
* fu_time_index: the follow-up time of the original paper


```{r index cohort, cache = TRUE}

tooth_index <- get_index_cohort(index_data = tooth,
                                mean_age_index = NULL,
                                sex_ratio_index = "women_only",
                                longitudinal_diseases = c("hd", "stroke", "lung",
                                                          "dia", "can", "dem"),
                                cross_sectional_diseases = c("anemia"),
                                fu_time_index = 6)

save(tooth_index_values, tooth_count, tooth_index,
     file = here::here("RESULTS_FINAL", "final_data", "tooth2008_data.RData"))
```

### Count cohort

Once you created the index cohort, you will then use it to create the count cohort. It will have the same follow up start the same follow up time and the same participants. But in this case, we will use all ten diseases included in the SHIFT data and count how many diseases each participant had at the start of the follow up. The function `get_count_cohort` creates this count cohort, based on the index cohort (the output of the `get_index_cohort` function).

The arguments of the function are:

* index_data: the index cohort dataset, output of the `get_index_cohort` function
* data: the original SHIFT data. By default, it is assumed to be the shift_data but it has be loaded into R beforehand

```{r count cohort, cache = TRUE}
tooth_count <- get_count_cohort(index_cohort_data = tooth_index)
```

Note that the sample size will likely be smaller than that for the index cohort as there is more variables in the count cohort and hence greater chance of any of the variables being missing. That's why we need to do one last step.

### Index cohort update

As the last step, we need to remove those patients who have complete follow up in the index cohort but who miss some variables (diseases) in the count cohort:


```{r index cohort update, cache = TRUE}
tooth_index <- tooth_index[tooth_index$ergoid %in% tooth_count$ergoid, ]
```

### Index values

The last step is to calculate the index value based on the assigned weights:

```{r index values, cache = TRUE}
tooth_index_values <- get_index_values(index_cohort_data = tooth_index,
                                      diseases = d,
                                      weights = w)
```

Now, we have two cohorts *tooth_index* and *tooth_count* that have the same participants, the same follow up start, the same follow-up time and the same outcome. We save them.

```{r save, cache = TRUE}
save(tooth_index_values, tooth_count, tooth_index,
     file = here::here("RESULTS_FINAL", "final_data", "tooth2008_data.RData"))
```

We can now move on to compare them. Hurrah :D

The workflow up until this point is executed in the *02_compile_index.R*.



## Data analysis

For data analysis, we use the `validate_index_v02` function. It has four arguments:

* dat: data containing both the index values and disease counts for a particular cohort
* index_name: name of the index to be used on plots and other output
* include_age: shall we include age as a separate independent variable? (Some indices incorporate age directly into their index value)
* include_sex" shall we include sex as a separate independent variable? (Some indices are sex specific)

This function fits a logistic regression model on the data and produce a series of metrics and outcomes:

* Basic information about the cohort (mean age, size, number of events, sex ratio)
* Performance of the count, index and base model in terms of Nagelkerke's R^2, C-statistics and Brier score
* Discrimination slope and discrimination box plots
* Net reclassification improvement
* Kaplan-Meier plots by age quartiles for index and count models
* Fits of the three models, as returned by the `rms::lrm` function
* calibration plots

```{r validate, cache = TRUE}
load(here::here("RESULTS_FINAL", "final_data", "desai2002_data.RData"))

df <- left_join(desai_count,
                desai_index_values |> dplyr::select(ergoid, index_value),
                by = "ergoid")


results_desai <- validate_index_v02(dat = df, index_name = "Desai 2002",
                                    include_age = TRUE)

save(results_desai,
     file = here::here("RESULTS_FINAL", "validation", "desai2002.RData"))
```

All data analysis is executed in the *03_validate_index.R* script.
